---
title: "Period-Doubling Bifurcation"
author: "Longxiao"
date: "2023-04-16"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: readable
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

> This is a demonstration of **period-doubling bifurcations**. Here, I
> demonstrate how period-doubling bifurcations emerge in a logistic
> population growth model as the intrinsic growth rate varies, using R
> for simulation. For the formulation of the simplified logistic model
> and the emergence of period-doubling bifurcations in population
> growth, refer to May’s classic paper [*Biological Populations with
> Nonoverlapping Generations: Stable Points, Stable Cycles, and
> Chaos*](https://www.science.org/doi/10.1126/science.186.4164.645).

In a dynamical system, the emergence of *chaos* is closely related to
**period-doubling bifurcations** that occur as system parameters change.
Simply put, as the parameter varies, the stable state of the system
bifurcates, eventually leading to chaos once a certain threshold is
reached. We begin with the simplest system that exhibits period-doubling
bifurcations, studying the dynamical characteristics of a difference
equation with a quadratic term, and attempt to understand the nature of
period-doubling bifurcations.

# Logistic Growth of a Population

The story begins with a classic question in theoretical biology and
ecology: How does a population change over time?

Consider a model where there is no generation overlap, and individuals
are born and die at constant rates $g$ and $b$, respectively. The
population grows at a constant rate $r = g - b$. A difference equation
with only a linear term (which behaves simply) is used to describe the
dynamics of such a model:

$$
N_t = r N_{t-1} \tag{1}
$$

We could simulate this system from $N_0 = 0.6$, with different intrinsic
rate $r = 0.6,-0.6,2,-2$ (include values that are positive or negative,
larger than $1$ or not, notice if $r = 1$ then $N_t = N_0$ for all $t$).

```{r}
Tt=15
r=c(0.6,-0.6,2,-2)
N0=0.6

SDF<-function(r,N0){
  N<-integer(Tt)
  N[1]=N0
  for(i in 2:Tt){
  N[i]=r*N[i-1]
  }
  return(N)
}

par(mfrow=c(2,2),pin=c(3,1))

for (j in r) {
  N<-SDF(j,N0)
  plot(c(1:Tt),N,type = "o")
  ptext<-paste("r=",j,"N0=",N0)
  abline(h=0,col="red")
  text(1,(600*abs(j)-359.5),pos= 4,ptext)
}
```

Or, we could simulate with different initial population size:

```{r}
Tt=15
r=2
N0=c(0.6,-0.6,2,-2)

par(mfrow=c(2,2),pin=c(3,1))

for (j in N0) {
  N<-SDF(r,j)
  plot(c(1:Tt),N,type = "o")
  ptext<-paste("r=",r,"N0=",j)
  abline(h=0,col="red")
  text(1,2000*j,pos= 4,ptext)
}
```

The system either diverges to infinity or converges to zero depending on
the initial state and the value of the growth rate parameter. When the
initial value and the growth rate have opposite signs, fluctuations may
occur, but the trend remains unchanged. Overall, the behavior is
monotonic and predictable.

With a continuous time, the system could also be modeled by a difference
equation:

$$
\frac{dN}{dt} = rN
$$

which has a solution $N_t = Ce^{rt}$. The differential equation looks
like ($r=2$, $N_0 = 0.01$, $t \in (1,8)$):

```{r}
r<-2
T1<-8
delta_t<-0.001
t<-seq(from = 0, to = T1, by = delta_t)
N0<-0.01

N<-integer(length(t)) #vector of population size
N[1]=N0       #initial population size N_0

for(i in 2:length(t)){
  N[i]=N[i-1]+delta_t*r*N[i-1]  #iterate function
}
  
logi_map<-cbind(t,N)
names(logi_map)<-c("t","N")
plot(logi_map, pch = ".")
```

However, a real population typically cannot grow at a constant rate, as
there are various limitations such as nutrition, space, and other
resources. As a result, we introduce an environmental capacity $K$ into
the model, which represents the maximum population size that can be
sustained by the environment:

$$
N_t = rN_{t-1}\left(1 - \frac{N_{t-1}}{K}\right) \tag{2}
$$

Here, $N_t$ represents the population size at time $t$, $r$ is the
intrinsic growth rate of the population, and $K$ is the environmental
capacity. With a simple transformation, the equation can be rewritten
into a general form for this class of models:

$$
N_t = rN_{t-1}(1 - N_{t-1}) \tag{3}
$$

Note that the system's dynamics are now governed by a single parameter
$r$.

> The transformation from Equation (2) to Equation (3) removes the
> dimensional units, making the dynamical properties clearer and more
> concise. (The specific transformation can be skipped.) $$
> \text{Let: } a_t = \frac{N_t}{K} \\
> a_t = ra_{t-1}(1 - a_{t-1}')
> $$
>
> which is the same form as Equation (3).

It is easy to see that under the given constraints: $N_t$ lies within
the interval [0, 1] (meaning the population size is positive and does
not exceed the environmental carrying capacity), and $r$ is greater than
zero (indicating population growth). Equation (3) has a fixed point:

$$
N^* = rN^*(1 - N^*)
$$

Solving for $a$ gives:

$$
N^* = 1 - \frac{1}{r} \tag{4}
$$

When the value of $r$ is relatively small, the system will converge to
the fixed point regardless of the initial value.

```{r}
Tt<-20
r<-2
N0<-c(0.01,0.3,0.5,0.8)
Nt<-length(N0)
N<-integer(Tt)

par(mfrow=c(2,2),pin=c(8,6))

for (j in N0) {
  N[1]=j
for(i in 2:Tt){
  N[i]=r*N[i-1]*(1-N[i-1])
}
t<-c(1:Tt)
par(pin=c(1.8,1.2))
plot(t,N,type = "o",ylim = c(0,1),yaxp = c(0,1,10))
abline(h=0.5,lwd=0.5,col="red")
ptext<-paste("N0=",j)
  text(0.5,0.8,pos= 4,ptext)
}

```

With smaller initial values $N_0$, the system exhibits behavior
characteristic of nonlinearity. Followed equation(4), the fixed point
should exist with all $r > 1$. But the system will not always converge
when the $r$ is getting large. We will discuss it later.

Under continuous-time conditions, this model is known as **logistic
growth**, described by the following equation:

$$
\frac{dN}{dt} = rN(1 - N)
$$

```{r}
r<-2
T1<-8
delta_t<-0.001
t<-seq(from = 0, to = T1, by = delta_t)
N0<-0.01

N<-integer(length(t)) #vector of population size
N[1]=N0       #initial population size N_0

for(i in 2:length(t)){
  N[i]=N[i-1]+delta_t*r*N[i-1]*(1-N[i-1])  #iterate function
}
  
logi_map<-cbind(t,N)
names(logi_map)<-c("t","N")
plot(logi_map, pch = ".")
```

# Period-Doubling Bifurcations

Now it is time to discuss the complex dynamics of the logistic model
with larger $r$. The exponential growth model has very simple behavior.
When we introduce a quadratic term into the system (as logistic model),
the situation seems not change significantly. However, as we gradually
increase the value of the growth parameter $r$, the state of the system
also changes accordingly:

We simulate the discrete-time logistic model with
$r = 0.8,2.5,3.2,3.5,3.55,3.6,3.99,4.1$

```{r}
Tt=50 #迭代次数
r=c(0.8,2.5,3.2,3.5,3.55,3.6,3.99,4.1) #增长率参数
N0=0.5 #初始值

N<-integer(Tt) #种群数量

par(mfrow=c(4,2))

for (j in r) {
  for(i in 2:Tt){
    N[1]=N0
    N[i]=j*N[i-1]*(1-N[i-1])
  }
  t<-c(1:Tt)
  par(pin=c(2.5,0.7))
  plot(t,N,type = "o",ylim = c(0,1),yaxp = c(0,1,10))
  abline(h=max((1-1/j),0),col="red")
  ptext<-paste("r=",j)
  text(0.5,0.1,pos= 4,ptext,col="red")
}
```

Although the system still tends to approach the fixed point in the long
run, it sometimes fails to reach it precisely and instead begins to
oscillate. And once the $r$ is getting much larger, it is even hard to
notice the oscillation. As the value of $r$ continues to increase, the
nature of the system's long-term fluctuations also changes.

It becomes evident that the period of these fluctuations varies with
$r$. In other words, the final state of the system—and how many such
states exist—depends on the value of $r$ (for example, $2$ when
$r = 3.2$, $4$ when $r = 3.5$). Also, we can notice that the number of
different state seems follows the series $n = 2^k$.

We can visualize the positions of the system's final states for
different values of $r$ in a diagram:

```{r}

logistic_ref<-function(r,N){
  return(r*N*(1-N))
} #logistic map
  
par_map<-function(r,N0,t,k,func){
  N<-integer(t) #population size 
  N[1]=N0       #initial population

  a0<-as.data.frame(matrix(nrow = 0, ncol = 2)) 
  #dataframe of r-N*

  for (j in r) {
    for(i in 2:t){
      N[i]=func(j,N[i-1])
    }
    a<-cbind(rep(j,k),N[(t-k+1):t])
    a0<-rbind(a0,a)
  }
  names(a0)<-c("r","N")
  return(a0)
}
```

```{r}
r=c(0.8,2.5,3.2,3.5,3.55,3.6,3.99,4.1) #r, intrinsic rate
N0=0.6 #initial population size
t=1000 #iterate times
k=100 #number of kept final state

ap1<-par_map(r,N0,t,k,logistic_ref)
plot(ap1,pch=20,cex = 1)

abline(v=r[1:7],col="red")

```

Next, we sample the parameter $r$ more densely and again plot the system
states. We can clearly observe the bifurcation of system states. The
system transitions from a single stable point, to oscillations between
two states, then four states, with increasingly longer periods, and
eventually enters an aperiodic chaotic regime.

```{r}
r=seq(from = 0.8, to = 4.1, by = 0.05)
N0=0.6
t=500
k=100

ap2<-par_map(r,N0,t,k,logistic_ref)
plot(ap2,pch=20,cex = 1)

```

```{r}
r=seq(from = 0.8, to = 4.1, by = 0.001)
N0=0.6
t=500
k=100

a0<-par_map(r,N0,t,k,logistic_ref)
plot(a0,pch = ".")
```

> Another way to generate a bifurcation diagram is by iterating from a
> single initial state. The transitions between different periodic
> behaviors appear rougher compared to the diagram above, which also
> depends on the resolution of the simulation.

```{r}
t<-200000
r<-seq(from=1.0, to=4.0, length.out=t)
N0<-0.01

N<-integer(t)
N[1]=N0
  
for (i in 2:t) {
    N[i]=r[i]*N[i-1]*(1-N[i-1])
  }
b<-cbind(r,N)
plot(b,pch = ".")
```

```{r}
t<-100000
r<-seq(from=0.5, to=4.0, length.out=t)
N0<-0.6

N<-integer(t) 
N[1]=N0       
  
for (i in 2:t) {
    N[i]=r[i]*N[i-1]*(1-N[i-1]) 
  }
b<-cbind(r,N)
plot(b,pch = ".")
```

# Basic Analysis

## Feigenbaum Constants: pattern of bifurcations

In period-doubling bifurcations, the period of oscillations doubles as
the parameter increases. We then search for the parameter values at
which state transitions occur, and investigate their underlying pattern.

```{r}
f <- 6 # Number of bifurcations to find
period <- 2 # Minimum period
cycle2 <- integer(f + 1)
t = 2000 # Number of iterations
N0 = 0.6 # Initial value

pointfind <- function(r, num_point) {
  # r is the logistic map parameter, num_point is the period size to search for
  
  N <- integer(t) # Population size
  N[1] = N0       # Initial population size
  
  for (i in 2:t) {
    N[i] = logistic_ref(r, N[i-1]) # Logistic mapping, iterating to get population size changes
  }
  
  if (abs(N[t] - N[t - num_point]) > 0.000001) return(1)
  # Check if it has a cycle, i.e., whether the population size is equal for the specified period
  # Since it is only approaching the limit cycle, the values do not need to be exactly equal, they just need to be very close
  # Return 1 if no cycle is found
  else return(0)
}

lr <- 0 # Left boundary
rr <- 4 # Right boundary

for (m in 0:f) {
  # m is the number of bifurcations on the left side, 2^m is the number of bifurcations on the left
  while ((rr - lr) > 0.0000001) {
    midr <- ((rr + lr) / 2) # Middle value of the current interval
    if (pointfind(midr, (period^m)) == 1) rr <- midr
    # If the cycle at the middle value is not the bifurcation left cycle, the middle value is in the bifurcation right side
    else lr <- midr
    # Otherwise, the middle value is on the bifurcation left side
  }
  cycle2[m + 1] <- midr # Found the bifurcation point
  rr <- 4 # Reset the right boundary to 4, left boundary remains the same
}
# Use binary search to find the bifurcation points

print(cycle2)
```

It seems that these values are converge to a value close to $3.57$,
which suggest that there is a limitation for period-doubling
bifurcations. Just like in the diagram, the red lines for period
doubling getting dense.

```{r}
plot(a0,pch = ".")
abline(v=1,lwd=0.5,col="red")
for (i in cycle2) {
  abline(v=i,lwd=0.5,col="red")
}
```

To find more patterns, we calculate the interval of period doubling:

```{r}
c2_gap<-integer(f+1)
c2_gap[1]<-cycle2[1]-1
for (i in 2:(f+1)) {
    c2_gap[i]<-cycle2[i]-cycle2[i-1]
}
print(c2_gap)

c2_gaprate<-integer(f)
for (i in 1:f) {
    c2_gaprate[i]<-c2_gap[i]/c2_gap[i+1]
}
print(c2_gaprate)
```

By roughly simulation and calculation, the value of parameter $r$ for
period doubling seems a geometric sequence, with common ratio around
$4.6$.

```{r}
fit<-lm(c2_gaprate ~ 1)
summary(fit)
```

We then fit the sequence to a linear model. The estimation of common
ratio is $4.6158\pm0.0539$. Actually, This value is called ***Feigenbaum
Constants***, $\delta$, which is a general constant in the world of
chaos, refers to [Quantitative universality for a class of nonlinear
transformations](https://doi.org/10.1007/BF01020332).

```{r}
boxplot(c2_gaprate)
F_cons<-4.669201609  #Feigenbaum Constants
abline(h = fit$coefficients,lwd=2, col="red")
abline(h = F_cons,lwd=2, col="blue")
```

Our estimation is close to Feigenbaum Constant. And based on the
Feigenbaum Constant, we could calculate the upper bound of $r$ for
period doubling area ($\sim 3.545$).

```{r}
c2_gap0<-2
c2_total0<-1+c2_gap0*(F_cons/(F_cons-1))
print(c2_total0)
c2_total1<-1+c2_gap[1]*(fit$coefficients/(fit$coefficients-1))
print(c2_total1)
```

```{r}
plot(a0,pch = ".")
abline(v=1,lwd=0.5,col="red")
for (i in cycle2) {
  abline(v=i,lwd=0.5,col="red")
}
abline(v=c2_total1,col="blue")
```

## Fractal Features

Each branch of the bifurcation shows a similar pattern, which is closely
related to fractal feature. Here we showed some parts of the above
period-doubling bifurcation.

```{r}
t=1000
N0=0.6
k=200
r_b1=seq(from = 2.95, to = 3.6, by = 0.0005)
map_b1<-par_map(r_b1,N0,t,k,logistic_ref)
plot(map_b1,pch = ".")
```

```{r}
t=1000 #迭代次数
N0=0.6 #初始值
k=200 #绘图保留次数
r1=seq(from = 3.62, to = 3.64, by = 0.00004) 
r2=seq(from = 3.73, to = 3.75, by = 0.00004) 
r3=seq(from = 3.82, to = 3.88, by = 0.00004) 

map1<-par_map(r1,N0,t,k,logistic_ref)
map2<-par_map(r2,N0,t,k,logistic_ref)
map3<-par_map(r3,N0,t,k,logistic_ref)
```

```{r}
layout(matrix(c(1,2,3),1,3,byrow = T),widths = c(1,1,2.4))
plot(map1,pch = ".")
plot(map2,pch = ".")
plot(map3,pch = ".")
```

## Chaos and Initial Condition Sensitivity

The upper bound of period bifurcation suggest chaos with larger $r$.
Again, we simulated the system with different $r$, but also with two
close initial state.

```{r}
Tt=20 #迭代次数
r=c(2.5,3.2,3.5,3.9) #增长率参数
N0=0.5 #初始值
l=0.01

N<-integer(Tt) #种群数量
N1<-integer(Tt)

par(mfrow=c(2,2))

for (j in r) {
  for(i in 2:Tt){
    N[1]=N0
    N1[1]<-N[1]-l
    N[i]=j*N[i-1]*(1-N[i-1])
    N1[i]=j*N1[i-1]*(1-N1[i-1])
  }
  t<-c(1:Tt)
  plot(t,N,type = "l",ylim = c(0,1),yaxp = c(0,1,10))
  lines(t,N1,col="red")
  ptext<-paste("r=",j)
  text(0.5,0.1,pos= 4,ptext,col="red")
}
```

As we can see here, when the $r$ is larger, the system dynamics is
sensitive with initial state, and the error becomes unpredictable. We
call this kind of phenomena ***chaos***.

```{r}
Tt=50
r=3.9
N0=0.5
l=0.001

N<-integer(Tt)
N1<-integer(Tt)
N2<-integer(Tt)

N[1]=N0
N1[1]<-N[1]-l
N2[1]<-N[1]-2*l
for(i in 2:Tt){
  N[i]=r*N[i-1]*(1-N[i-1])
  N1[i]=r*N1[i-1]*(1-N1[i-1])
  N2[i]=r*N2[i-1]*(1-N2[i-1])
}
t<-c(1:Tt)

plot(t,N,type = "l",col="green",lwd = 2,ylim = c(0,1),yaxp = c(0,1,10),)
lines(t,N1,col="blue",lwd = 2)
lines(t,N2,lwd = 2)

ptext<-paste("r=",j)
text(0.5,0.05,pos= 4,ptext,col="red")
```

# Another example

Not only the logistic map have such an amazing property. Actually, other
nonlinear functions could also work. For example, the sine function
mapping:

```{r}
sin_map<-function(r,N){
  return(r*sin(N))
} #logistic map

r=seq(from = 0.8, to = 4.1, by = 0.001)
N0=0.6
t=500
k=100

b0<-par_map(r,N0,t,k,sin_map)
plot(b0,pch = ".")
```

```{r}
f <- 6 # Number of bifurcations to find
period <- 2 # Minimum period
cycle3 <- integer(f + 1)
t = 2000 # Number of iterations
N0 = 0.6 # Initial value

pointfind <- function(r, num_point) {
  # r is the logistic map parameter, num_point is the period size to search for
  
  N <- integer(t) # Population size
  N[1] = N0       # Initial population size
  
  for (i in 2:t) {
    N[i] = sin_map(r, N[i-1]) # Logistic mapping, iterating to get population size changes
  }
  
  if (abs(N[t] - N[t - num_point]) > 0.000001) return(1)
  # Check if it has a cycle, i.e., whether the population size is equal for the specified period
  # Since it is only approaching the limit cycle, the values do not need to be exactly equal, they just need to be very close
  # Return 1 if no cycle is found
  else return(0)
}

lr <- 0 # Left boundary
rr <- 4 # Right boundary

for (m in 0:f) {
  # m is the number of bifurcations on the left side, 2^m is the number of bifurcations on the left
  while ((rr - lr) > 0.0000001) {
    midr <- ((rr + lr) / 2) # Middle value of the current interval
    if (pointfind(midr, (period^m)) == 1) rr <- midr
    # If the cycle at the middle value is not the bifurcation left cycle, the middle value is in the bifurcation right side
    else lr <- midr
    # Otherwise, the middle value is on the bifurcation left side
  }
  cycle3[m + 1] <- midr # Found the bifurcation point
  rr <- 4 # Reset the right boundary to 4, left boundary remains the same
}
# Use binary search to find the bifurcation points

print(cycle3)
```

```{r}
c3_gap<-integer(f+1)
c3_gap[1]<-cycle3[1]-1
for (i in 2:(f+1)) {
    c3_gap[i]<-cycle3[i]-cycle3[i-1]
}

c3_gaprate<-integer(f)
for (i in 1:f) {
    c3_gaprate[i]<-c3_gap[i]/c3_gap[i+1]
}
print(c3_gaprate)
```

```{r}
fit<-lm(c3_gaprate ~ 1)
summary(fit)
```

And we can also draw Feigenbaum Constant.
