---
title: "Brief Introduction to PCA"
author: "Longxiao"
date: "2025-01-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here we use a simulated dataset to illustrate the process of Principal Component Analysis (PCA). The original data is three-dimensional, and it was simulated based on a plane described by $z = 2x+3y$, with some added random noise. As a result, this dataset can be assumed to lie in a two-dimensional space, with two independent vectors defining the surface.

```{r, warning=FALSE}
# Load necessary libraries
library(ggplot2)
library(scatterplot3d)

# Set the seed for random number generation
set.seed(123)

# Number of samples (data points) to generate
n_samples <- 20

# Generate n_samples random numbers for x and y in the range [-1, 1] from a uniform distribution
x <- runif(n_samples, -1, 1)
y <- runif(n_samples, -1, 1)

# Define a plane equation z = 2 * x + 3 * y

# z0 represents the ideal values for z without any noise
z0 <- 2 * x + 3 * y 
# Generate random noise with mean 0 and standard deviation 1
z <- z0 + rnorm(n_samples, mean = 0, sd = 1)

# Combine x, y, and z
data_3D <- data.frame(x, y, z)

head(data_3D)
```

By visualizing the data, we can observe that the 20 data points are roughly distributed around the plane.

```{r}
s3d <- scatterplot3d(data_3D$x, data_3D$y, data_3D$z, 
              color = "blue", pch = 16, 
              main = "Original 3D Data", 
              xlab = "X Axis", ylab = "Y Axis", zlab = "Z Axis")

s3d$plane3d(Intercept = 0, x.coef = 2, y.coef = 3, col = "lightblue", lty = 2, lwd = 1.2)
```

Then, by PCA analysis, the data is reduced to two dimensions.

```{r}
pca_result <- prcomp(data_3D, center = TRUE, scale. = FALSE)

data_2D <- as.data.frame(pca_result$x[, 1:2])
colnames(data_2D) <- c("PC1", "PC2")

ggplot(data_2D, aes(x = PC1, y = PC2)) +
  geom_point(color = "red", alpha = 2) +
  ggtitle("PCA Reduced 2D Data") +
  xlab("Principal Component 1") +
  ylab("Principal Component 2") +
  theme_minimal()
```

but...

```{r}
data0 <- data.frame(x, y, z0)

# Define two vectors (v_x and v_y) that lie on the plane in 3D space.
v_x <- c(-0.1, 0, -0.2)
v_y <- c(0, -0.1, -0.3)

# Create a transformation matrix by the two vectors v_x and v_y
transformation_matrix <- cbind(v_x, v_y)[1:2, ]

# Apply the linear transformation to the original data
transformed_data <- data.frame(
  x_transformed = data0$x * v_x[1] + data0$y * v_y[1],
  y_transformed = data0$x * v_x[2] + data0$y * v_y[2] )

ggplot(transformed_data, aes(x = x_transformed, y = y_transformed)) +
  geom_point(color = "blue") +
  labs(title = "2D Visualization of Data after Linear Transformation", x = "Transformed X", y = "Transformed Y") +
  theme_minimal()
```

they are quite differnet!

We perform the PCA step by step.

## 1. standardization

$$
Z_{ij} = \frac{X_{ij} - \mu_{j}}{\sigma_{j}}
$$

where $i$ is the order of data, $j$ is the dimension.

```{r}
data_standardized <- scale(data_3D)
head(data_standardized)
```

## 2. covariance matrix

$$
Cov = \frac{1}{n-1} X^T X
$$

```{r}
cov_matrix <- cov(data_standardized)

cov_matrix
```

## 3. eigenvalue and eigenvector of the cov matrix

$$
Cv = \lambda v
$$

```{r}
eigen_result <- eigen(cov_matrix)
eigenvalues <- eigen_result$values
eigenvectors <- eigen_result$vectors

eigen_result
```

## 4. choose dimensions

```{r}
sorted_indices <- order(eigenvalues, decreasing = TRUE)
top_2_eigenvectors <- eigenvectors[, sorted_indices[1:2]]

sorted_indices
```

Here are the two eigenvectors we selected (red and green), with a set of independent vectors that lie on the plane (dark red $(1,0,2)$ and dark green $(0.6,-0.5,-0.3)$).

```{r}
s3d <- scatterplot3d(data_3D$x, data_3D$y, data_3D$z,
                     color = "blue", pch = 16, 
                     main = "3D Data with PCA Components",
                     xlab = "X Axis", ylab = "Y Axis", zlab = "Z Axis")

s3d$plane3d(Intercept = 0, x.coef = 2, y.coef = 3, col = "lightblue", lty = 2, lwd = 1.2)

s3d$points3d(c(0, top_2_eigenvectors[1, 1]), 
             c(0, top_2_eigenvectors[2, 1]), 
             c(0, top_2_eigenvectors[3, 1]), 
             col = "red", type = "l", lwd = 2)

s3d$points3d(c(0, top_2_eigenvectors[1, 2]), 
             c(0, top_2_eigenvectors[2, 2]), 
             c(0, top_2_eigenvectors[3, 2]), 
             col = "green", type = "l", lwd = 2)

s3d$points3d(c(0, 0.6), 
             c(0, 0), 
             c(0, 1.2), 
             col = "darkred", type = "l", lwd = 5)

s3d$points3d(c(0, 0.6), 
             c(0, -0.5), 
             c(0, 0.3), 
             col = "darkgreen", type = "l", lwd = 5)

s3d$points3d(data_3D$x, data_3D$y, data_3D$z, col = "blue", pch = 16)
```

## 5. Mapping the data to the new space

```{r}
pca_result <- data_standardized %*% top_2_eigenvectors

head(pca_result)
```

```{r}
data_2D <- as.data.frame(pca_result[, 1:2])
colnames(data_2D) <- c("PC1", "PC2")

ggplot(data_2D, aes(x = PC1, y = PC2)) +
  geom_point(color = "red", alpha = 5) +
  ggtitle("PCA Reduced 2D Data") +
  xlab("Principal Component 1") +
  ylab("Principal Component 2") +
  theme_minimal()
```

Here we could explore more about the sample size and data noise (which is the variance $\sigma$ here).

```{r}
pca_test<-function(n,sigma){
  n_samples <- n
  x <- runif(n_samples, -1, 1)  
  y <- runif(n_samples, -1, 1)
  z0 <- 2 * x + 3 * y 
  z <- z0 + rnorm(n_samples, mean = 0, sd = sigma)
  data_3D <- data.frame(x, y, z)
  
  v_x <- c(-0.1, 0, -0.2)
  v_y <- c(0, -0.1, -0.3)
  
  data_standardized <- scale(data_3D)
  cov_matrix <- cov(data_standardized)
  eigen_result <- eigen(cov_matrix)
  eigenvalues <- eigen_result$values
  eigenvectors <- eigen_result$vectors
  
  sorted_indices <- order(eigenvalues, decreasing = TRUE)
  top_2_eigenvectors <- eigenvectors[, sorted_indices[1:2]]
  
  s3d <- scatterplot3d(data_3D$x, data_3D$y, data_3D$z,
                     color = "blue", pch = 16, 
                     main = "3D Data with PCA Components",
                     xlab = "X Axis", ylab = "Y Axis", zlab = "Z Axis", type = "n")

s3d$plane3d(Intercept = 0, x.coef = 2, y.coef = 3, col = "lightblue", lty = 2, lwd = 1.2)

s3d$points3d(c(0, top_2_eigenvectors[1, 1]), 
             c(0, top_2_eigenvectors[2, 1]), 
             c(0, top_2_eigenvectors[3, 1]), 
             col = "red", type = "l", lwd = 2)

s3d$points3d(c(0, top_2_eigenvectors[1, 2]), 
             c(0, top_2_eigenvectors[2, 2]), 
             c(0, top_2_eigenvectors[3, 2]), 
             col = "green", type = "l", lwd = 2)

s3d$points3d(c(0, 0.6), 
             c(0, 0), 
             c(0, 1.2), 
             col = "darkred", type = "l", lwd = 5)

s3d$points3d(c(0, 0.6), 
             c(0, -0.5), 
             c(0, -0.3), 
             col = "darkgreen", type = "l", lwd = 5)

s3d$points3d(data_3D$x, data_3D$y, data_3D$z, col = "blue", pch = 16, type = "n")
}
```

```{r}
#set.seed(2)
pca_test(10000,1)
```

```{r}
#set.seed(2)
pca_test(10000,1)
```

```{r}
pca_test(10000,1)
```

```{r}
pca_test(10,1)
```

```{r}
pca_test(10,1)
```

```{r}
pca_test(10,1)
```
