---
title: "Analysis of Spatial Coexistence and Neighbor Relationships in Forest Using Planar Graph Model"
author: "Longxiao"
date: "2021-04-21"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: readable
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In the study of organisms that live in a sessile manner (such as various
plants, corals, and sedentary mollusks), direct observations of their
interaction behaviors are difficult. Therefore, the interaction
relationships are often inferred from their distribution or spatial
coexistence. For example, it is generally inferred that species
exhibiting low co-occurrence tend to have strong competitive
interactions, whereas those with frequent co-occurrence are likely to
experience weak competition or engage in mutualistic interactions. In
the case of woody plants, spatial coexistence typically considers the
neighbor relationships between individual plants. However, previous
studies have generally described the neighbors of an individual as all
other individuals within a certain spatial range (Hubbell et al., 2001;
Peters, 2003; Stoll and Newbery, 2005; Comita et al., 2010). This method
is computationally simple and requires fewer resources, but its
definition of neighbors is vague, and the selection of range is often
based on experience rather than a rigorous mathematical foundation.

To better define and analyze the spatial coexistence relationships of
woody plants, we developed a method based on planar graphs to analyze
neighbor relationships.

```{r, warning=FALSE, message=FALSE}
#packages
library(ggplot2)
library(pheatmap)
library(dplyr)
library(deldir)
```

# Defining Neighbor Relationships in a Planar Graph

Consider a set of points in a finite plane, which represent the
individuals in forest.

```{r}
# generate example: poisson distributed random points in 1*1 plane, set density as \rho.
example_generator<-function(rho = 100, X = 1, Y = 1, seed = 123){
  set.seed(seed)
  # possion distributed points
  n_poisson <- rpois(1, lambda = rho * X * Y)
  poisson_points <- data.frame(
    x = runif(n_poisson, 0, X),
    y = runif(n_poisson, 0, Y),
    type = "sample"
    )
  return(poisson_points)
}
```

```{r}
rho = 100
X = 1
Y = 1
seed = 123

tree_sample <- example_generator()

#visualization
p1 <- ggplot(tree_sample, aes(x, y)) +
  geom_point(color = "blue", alpha = 0.6) +
  coord_fixed() +
  ggtitle("simulated tree individuals in a spot")+
  theme_minimal()

p1
```

We propose the following model: each individual is associated with a
neighborhood centered on itself, with neighborhoods of different
individuals being non-overlapping. Furthermore, the union of all these
neighborhoods completely covers the sampling area. In other words, these
neighborhoods form a partition of the sampling region.

```{r, echo=FALSE}
voronoi <- deldir(tree_sample$x, tree_sample$y)

voronoi_edges <- voronoi$dirsgs

ggplot() +
  geom_segment(data = voronoi_edges, 
               aes(x = x1, y = y1, xend = x2, yend = y2), 
               color = "darkgreen", alpha = 0.6) +
  geom_point(data = tree_sample, aes(x = x, y = y), color = "blue", alpha = 0.6) +
  coord_fixed() +
  theme_minimal() +
  ggtitle("Neighborhoods of Sample Points")
```

Neighbor relationships are defined by the spatial proximity of the
neighborhoods. Specifically, if the neighborhoods of two points are
adjacent, these two points are considered neighbors. In this model, the
neighbor relationship can be represented as a set of edges in the plane,
where these edges do not intersect. In other words, the total neighbor
relationships form a planar graph.

# Complete Occupation in Subtropical Forests

Considering that subtropical forests often have a high canopy density,
we may assume that the occupation of the plane by the set of points is
sufficiently complete. From this, we infer that the number of neighbor
relationships between points should be maximized, ensuring that there
are no disconnected edges on the plane that would meet the required
conditions. This leads to the generation of a complete planar graph.

# Minimizing the Spatial Distance Between Neighboring Points

Since we are focusing on spatial neighbor relationships, the goal is to
find the nearest set of neighboring points such that the spatial
distance between neighbors is minimized. Additionally, for any given
point, its neighbors must form a polygon containing only that point.
This necessitates the creation of a minimum complete planar graph.

# Mathematical Constraints for the Planar Graph $G(V, E)$

As state above, the graph $G(V, E)$ must satisfy the following
mathematical constraints:

1.  **Non-intersecting Edges:** $$
    \forall e_i, e_j \in E, e_i \cap e_j = \emptyset
    $$ This means that the edges $e_i$ and $e_j$ should not intersect.

2.  **Minimization of Edge Length:** $$
    \min \sum e_i \quad \text{for} \quad e_i \in E
    $$ This minimizes the total length of the edges.

3.  **Maximization of Vertex Degree:** $$
    \max \sum \text{degree}(v_i) \quad \text{for} \quad v_i \in V
    $$ This maximizes the degree of the vertices (i.e., the number of
    edges connected to a vertex).

Where: - $V$ and $E$ represent the set of vertices and edges of the
graph $G$, - $\text{degree}(v_i)$ denotes the degree of the vertex
$v_i$, which is the number of edges connected to that vertex.

# The Algorithm Used to Implement the Model

We implement this process in real-world woody plant distribution data
(here illustrated with the example generated above) using a global
search approach. The process begins by generating a set of boundary
points that define the sampling plot boundaries. This avoids incorrect
neighbor relationships that might be formed due to boundary constraints.

```{r}
boundary_generator <- function(rho = 100, X = 1, Y = 1){
  # average distance scale
  avg_dist <- sqrt(X * Y / rho)
  
  # boundary area
  x_seq <- seq(-avg_dist, X + avg_dist, by = avg_dist)
  y_seq <- seq(-avg_dist, Y + avg_dist, by = avg_dist)
  
  boundary_x <- expand.grid(
    x = x_seq, y = c(-avg_dist, Y + avg_dist))
  boundary_y <- expand.grid(
    x = c(-avg_dist, X + avg_dist), y = y_seq)
  
  boundary_points <- rbind(boundary_x, boundary_y)
  boundary_points <- unique(boundary_points)
  boundary_points$type <- "boundary"
  
  return(boundary_points)
}
```

```{r}
boundary <- boundary_generator()
sample_data <- rbind(tree_sample,boundary)

p2 <- ggplot(sample_data, aes(x, y, color = type)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = c("boundary" = "red", "sample" = "blue"))+
  coord_fixed() +
  ggtitle("Sample Points and Boundary Points") +
  theme_minimal()

p2
```

```{r,echo=FALSE}
voronoi <- deldir(sample_data$x, sample_data$y)

voronoi_edges <- voronoi$dirsgs

ggplot() +
  geom_segment(data = voronoi_edges, 
               aes(x = x1, y = y1, xend = x2, yend = y2), 
               color = "darkgreen", alpha = 0.6) +
  geom_point(data = sample_data, aes(x = x, y = y, color = type), alpha = 0.6) +
  scale_color_manual(values = c("boundary" = "red", "sample" = "blue"))+
  coord_fixed() +
  theme_minimal() +
  ggtitle("Neighborhoods with boundary points")
```

Using the individual location data, we compute the pairwise distances
between all individuals in the sampling plot, thus constructing a
distance matrix. This matrix represents a complete graph for the set of
points, where the edges include all potential connections.

```{r}
#calculate distance matrix
dist_generator <- function(all_points){
  #distance matrix
  dist_matrix <- as.matrix(dist(all_points[, c("x", "y")]))
  #lower part
  lower_idx <- which(lower.tri(dist_matrix), arr.ind = TRUE)
  
  #translate to a dataframe
  distance_df <- data.frame(
    i = lower_idx[, 1],
    j = lower_idx[, 2],
    distance = dist_matrix[lower_idx]
    )
  return(distance_df)
}
```

```{r}
distance_df <- dist_generator(sample_data)
ggplot(distance_df, aes(x = i, y = j, fill = distance)) +
  geom_tile() +
  scale_fill_viridis_c(option = "magma") +
  coord_fixed() +
  theme_minimal() +
  ggtitle("Distance Matrix Heatmap")
```

Next, we create a set of neighbor relationships. To reduce computational
complexity, we filter the edges within a specific distance range (for
example, 3 or 4 times of the average distance) and sort them by
distance.

```{r}
# sort and preprocess distance dataframe
dist_sort <- function(distance_df, max_dist){
  distance_df <- distance_df %>%
  filter(distance < max_dist) %>%
  arrange(distance)
  return(distance_df)
}
```

```{r}
avg_dist <- sqrt(X * Y / rho)
distance_df_sort <- dist_sort(distance_df, 4*avg_dist)

ggplot(distance_df_sort, aes(x = i, y = j, fill = distance)) +
  geom_tile() +
  scale_fill_viridis_c(option = "magma") +
  coord_fixed() +
  theme_minimal() +
  ggtitle("Distance Matrix Heatmap")
```

We then check each edge from smallest to largest to see if it intersects
with any existing edges in the current neighbor relationship set. If it
does not intersect, it is added to the set. This process continues until
the complete set of neighbor relationships is generated.

```{r}
# examine intersect with both end of the edge
is_intersect <- function(p1, p2, q1, q2) {
  cross <- function(p, q, r) {
    (q$x - p$x) * (r$y - p$y) - (q$y - p$y) * (r$x - p$x)
  }
  
  (cross(p1, p2, q1) * cross(p1, p2, q2) < 0) &&
  (cross(q1, q2, p1) * cross(q1, q2, p2) < 0)
}
```

```{r}
neighbor_generator <- function(all_points, distance_df_sort){
  
  selected_edges <- data.frame(i = integer(), j = integer(), distance = double())

  for (k in 1:nrow(distance_df_sort)) {
    i <- distance_df_sort$i[k]
    j <- distance_df_sort$j[k]
    d <- distance_df_sort$distance[k]
  
    p1 <- all_points[i, c("x", "y")]
    p2 <- all_points[j, c("x", "y")]
  
    has_intersection <- FALSE
    if (nrow(selected_edges) > 0) {
      for (l in 1:nrow(selected_edges)) {
        i_exist <- selected_edges$i[l]
        j_exist <- selected_edges$j[l]
      
        q1 <- all_points[i_exist, c("x", "y")]
        q2 <- all_points[j_exist, c("x", "y")]
      
        if (is_intersect(p1, p2, q1, q2)) {
          has_intersection <- TRUE
          break
        }
      }
    }
  
    if (d == 0 || !has_intersection) {
      selected_edges <- rbind(selected_edges, data.frame(i = i, j = j, distance = d))
    }
  }
  
  return(selected_edges)
}
```

```{r}
edges_data <- neighbor_generator(sample_data,distance_df_sort)
head(edges_data)
```

```{r}
# both ends of the edges
edges_ends <- edges_data %>%
  rowwise() %>%
  mutate(x1 = sample_data$x[i], y1 = sample_data$y[i], x2 = sample_data$x[j], y2 = sample_data$y[j])

p3<-ggplot() +
  geom_segment(data = edges_ends, aes(x = x1, y = y1, xend = x2, yend = y2), color = "darkgray", linewidth = 0.5, alpha = 0.7) +
  geom_point(data = sample_data, aes(x = x, y = y, color = type), alpha = 0.6) +
  scale_color_manual(values = c("boundary" = "red", "sample" = "blue"))+
  ggtitle("Final Network of Selected Segments") +
  theme_minimal() +
  coord_fixed()

p3
```

Then, we delete boundary points and edges related to boundary, the final
result contains all coexistence pairs.

```{r}
# delete boundary points
edges_no_boundary <- edges_data %>%
  filter(i %in% rownames(tree_sample) & j %in% rownames(tree_sample))

edges_end_no_boundary <- edges_no_boundary %>%
  rowwise() %>%
  mutate(x1 = tree_sample$x[as.numeric(i)], y1 = tree_sample$y[as.numeric(i)],
         x2 = tree_sample$x[as.numeric(j)], y2 = tree_sample$y[as.numeric(j)])

p4<-ggplot() +
  geom_segment(data = edges_end_no_boundary, aes(x = x1, y = y1, xend = x2, yend = y2), color = "darkgray", linewidth = 0.5, alpha = 0.7) +
  geom_point(data = tree_sample, aes(x = x, y = y), color = "blue", size = 2, alpha = 0.7) +
  ggtitle("Network") +
  theme_minimal() +
  coord_fixed()

p4
```

And we could inferring the interaction strength as well as interaction
networks accordingly.

# Discussion

The algorithm we implemented here follows a greedy approach. However,
when the number of individuals increases significantly, this global
algorithm becomes inefficient and time-consuming. To enhance its
performance, several improvements can be considered:

-   Divide the area into smaller patches: This can help by reducing the
    problem size in each patch, making the computation more manageable.

-   Utilize spatial indexing techniques: When the area is extremely
    large, spatial index methods can speed up the process by quickly
    identifying relevant regions.

-   Limit the number of final edges: By setting a constraint on the
    maximum number of edges, we can avoid considering distant edges that
    are unnecessary. In a complete planar graph, it can be proven that
    the number of edges is bounded by $3n−3$, where $n$ is the number of
    points.

Additionally, another more efficient approach is Delaunay triangulation,
which is significantly faster than the method we used here. Delaunay
triangulation represents the dual graph of a Voronoi diagram and
provides a more computationally efficient way to connect the points.

```{r}
delaunay_result <- deldir(sample_data$x, sample_data$y)

delaunay_edges <- delaunay_result$delsgs

ggplot() +
  geom_segment(data = delaunay_edges, 
               aes(x = x1, y = y1, xend = x2, yend = y2), 
               color = "darkgray", linewidth = 0.5) +
  geom_point(data = sample_data, aes(x = x, y = y, color = type), alpha = 0.6) +
  scale_color_manual(values = c("boundary" = "red", "sample" = "blue"))+
  coord_fixed() +
  theme_minimal() +
  ggtitle("Delaunay Triangulation")
```

> I became acquainted with the Voronoi graph only after completing this
> work; otherwise, I would have opted to use it from the outset.
> Delaunay triangulation, however, doesn’t seem to perfectly align with
> our model in some respects. While I have not yet fully clarified the
> exact differences between them, it is clear that they sometimes
> produce different edges.
>
> Additionally, (to make this work appear at least somewhat meaningful),
> the algorithm we developed can be easily extended to accommodate
> various other conditions. For instance, if we assume that the area of
> an individual’s neighborhood is related to its biomass (which could be
> measured by height or DBH), we can adapt our method accordingly. In
> such cases, the only modification needed would be to adjust how we
> calculate the distance, such as by dividing it by the height.

# Previous Code (by 2022)

```{r,eval=FALSE}
install.packages("dplyr")
install.packages("ggplot2")
install.packages("data.table")
```

```{r,eval=FALSE}
# Construct neighbor web of trees 
# By Chang Longxiao, 2019 

# Read data, including real data and edges that are chosen
data.final <- read.table("data.final.csv", header = TRUE, sep = ",", encoding = "UTF-8") 
data.edge <- read.table("edge.final.csv", header = TRUE, sep = ",", encoding = "UTF-8") 
names(data.final) <- c("I", "sp", "x", "y", "h", "DBH") 
names(data.edge) <- c("I", "sp", "x", "y", "h", "DBH") 

n <- length(table(data.final$I)) 

# Divide data by site 
data.list <- list() 
for (i in 1:n) { 
  data.sub <- subset(data.final, I == i) 
  data.edge$h <- mean(data.sub$h) 
  data.edge$DBH <- mean(data.sub$DBH) 
  data.list[[i]] <- rbind(data.sub, data.edge) 
} 

# Generate the web and store by site 
result.list <- list() 
for (i in 1:n) { 
  a <- as.data.frame(data.list[i])  
  result.list[[i]] <- neibor.calculate(a, 5, weight = F) 
} 

# Function to generate neighbor network 
neibor.calculate <- function(tree.plot, gapl, weight) { 
  num <- length(tree.plot[, 1]) 
  num.count <- length(tree.plot[tree.plot$I != 0, 1]) 
  m.dis <- matrix(0, num, num) 
  for (i in 1:num) { 
    for (j in 1:num) { 
      m.dis[i, j] <- sqrt((tree.plot$x[i] - tree.plot$x[j])^2 + (tree.plot$y[i] - tree.plot$y[j])^2) 
      if (weight) 
        m.dis[i, j] <- m.dis[i, j] / (tree.plot$DBH[i] + tree.plot$DBH[j]) 
    } 
  } 
  
  # Structure distance matrix
  exis <- matrix(0, num, num) 
  a <- 0 
  for (i in 1:(num - 1)) { 
    for (j in (i + 1):num) { 
      if (m.dis[i, j] <= gapl) a <- a + 1 
      else exis[i, j] <- -1 
    } 
  } 
  
  # Distances that match the gap 
  while (a > 0) { 
    p <- 1 
    q <- 2 
    l <- gapl 
    for (i in 1:(num - 1)) { 
      for (j in (i + 1):num) { 
        if (exis[i, j] == 0 & m.dis[i, j] < l) { 
          p <- i 
          q <- j 
          l <- m.dis[i, j] 
        } 
      } 
    } 
    print(l) 
    
    # Min distance now 
    if (l == 0) exis[p, q] <- 1 
    if (l > 0) { 
      for (i in 1:(num - 1)) { 
        if (i == p) next 
        for (j in (i + 1):num) { 
          if (j == q) next 
          if (exis[i, j] == 1 & m.dis[i, j] > 0) { 
            ij1 <- (tree.plot$x[j] - tree.plot$x[i]) * (tree.plot$y[p] - tree.plot$y[i]) - 
              (tree.plot$y[j] - tree.plot$y[i]) * (tree.plot$x[p] - tree.plot$x[i]) 
            ij2 <- (tree.plot$x[j] - tree.plot$x[i]) * (tree.plot$y[q] - tree.plot$y[i]) - 
              (tree.plot$y[j] - tree.plot$y[i]) * (tree.plot$x[q] - tree.plot$x[i]) 
            pq1 <- (tree.plot$x[q] - tree.plot$x[p]) * (tree.plot$y[i] - tree.plot$y[p]) - 
              (tree.plot$y[q] - tree.plot$y[p]) * (tree.plot$x[i] - tree.plot$x[p]) 
            pq2 <- (tree.plot$x[q] - tree.plot$x[p]) * (tree.plot$y[j] - tree.plot$y[p]) - 
              (tree.plot$y[q] - tree.plot$y[p]) * (tree.plot$x[j] - tree.plot$x[p]) 
            if (is.na(ij1 * ij2 < 0 & pq1 * pq2 < 0)) { 
              exis[p, q] <- -1 
              break 
            } 
            if (ij1 * ij2 < 0 & pq1 * pq2 < 0) { 
              exis[p, q] <- -1 
              break 
            }  
          } 
        } 
        if (exis[p, q] == -1) break 
      } 
      if (exis[p, q] == 0) exis[p, q] <- 1 
    } 
    a <- a - 1 
    print(a) 
  } 
  
  # Find edges of the graph 
  for (i in 2:num) { 
    for (j in 1:(i - 1)) { 
      if (exis[j, i] == -1) exis[j, i] <- 0 
      exis[i, j] <- exis[j, i] 
    } 
  } 
  
  # Fill the matrix 
  result <- list(m.dis = m.dis[1:num.count, 1:num.count], exis = exis[1:num.count, 1:num.count]) 
  return(result) 
} 

# Translate the neighborhood matrix into co-occurrence matrix 
inter.list <- list() 
for (i in 1:n) { 
  inter.list[[i]] <- inter.frame(data.list[[i]], result.list[[i]]$exis, 45) 
} 

# Function to create the co-occurrence matrix 
inter.frame <- function(sp.data, exis, sp.num) { 
  inter.num <- sum(exis) / 2 
  i.frame <- as.data.frame(matrix(0, inter.num, sp.num)) 
  n <- length(sp.data[, 1]) - 88 
  flag <- 1 
  for (i in 1:(n - 1)) { 
    for (j in (i + 1):n) { 
      if (exis[i, j] == 1) { 
        i.frame[flag, sp.data$sp[i]] <- 1 
        i.frame[flag, sp.data$sp[j]] <- 1 
        flag <- flag + 1 
      } 
    } 
  } 
  return(i.frame) 
}
```
